---
title: "Have Hurricanes Gotten More Severe"
output:
  html_notebook: default
  html_document:
    theme: null
    highlight: null
  pdf_document: default
bibliography: hurricane_severity.bibtex
---

```{r, message=FALSE, warning=FALSE, echo=FALSE}
source('db_conn.R')
source('hurricanes.R')
conn <- hurr_db_connect()

library(dplyr)
library(forecast)
library(leaflet)
library(ggplot2)
library(pander)
# library(rjson) Trash
```

```{r ggplot Default Theme, echo=FALSE}
theme_set(theme_bw() + theme(
    legend.position="bottom",
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(colour = "black")))
```

## Have Hurricanes Become Worse?
When Hurricane Katrina made landfall over Louisiana, it quickly became one of the worst natural disasters in American history.
This led many to single out Katrina as the result of global warming. However, "there is no way to prove that Katrina either was, or was not, affected by global warming. For a single event, regardless of how extreme, such attribution is fundamentally impossible. We only have one Earth, and it will follow only one of an infinite number of possible weather sequences." [@real_climate_katrina]

Even without climate change, severe storms like Katrina can still happen due to pure chance. However, we can still ask if *the probability of severe storms* has 
increased in recent years.

### Measuring Hurricane Intensity: Saffir-Simpson Hurricane Scale

```{r "SSHS Diagram"}
sshs_diagram()
```

### Frequency and Intensity of Storms
Before you freak out at the graph below, note that important pieces of weather tracking.

```{sql connection=conn, output.var='hurr_by_year'}
SELECT
  season, count(name), round(AVG(wind_max_kt)::numeric, 1),        
  saffir_simpson(AVG(wind_max_kt))
FROM hurricane_summary
WHERE season > 1900 AND wind_max_kt::int != -999
GROUP BY season
ORDER BY season
```

```{r "Hurricanes per Season", echo=FALSE}
add_notes(ggplot(hurr_by_year, aes(x=season, y=count, fill=saffir_simpson)) +
  geom_bar(stat='identity') +
  
  # Rescale to fit text annotations
  scale_y_continuous(limits=c(0, 160)) + 
  
  # Manually set legend for hurricane intensity
  scale_fill_manual(
    "Average Intensity",
    values = c("Tropical Depression" = "#5ebaff",
               "Tropical Storm" = "#00faf4",
               "Category 1" = "#ffffcc")
  ) +
  labs(
    title="Number of Hurricanes per Season",
    x="Year",
    y="Number"
  ))
```

From the figure above, there doesn't appear to be any change in the frequency of hurricanes since 1970.
For climate scientists, this is unsurprising because most hurricane models--when considering global warming,
while predicting stronger storms, do not predict more storms. (cite)

### Have Severe Hurricanes Become More Common
```{sql connection=conn, output.var='hurr_by_cat_year'}
SELECT
  season, saffir_simpson(wind_max_kt), count(name)
FROM hurricane_summary
WHERE season > 1900 AND wind_max_kt::int != -999
GROUP BY saffir_simpson(wind_max_kt), season
ORDER BY season
```

```{r echo=FALSE}
add_notes(ggplot(hurr_by_cat_year, aes(x=season, y=count, fill=saffir_simpson)) +
  geom_bar(stat='identity') +
  scale_fill_manual(
    "Intensity (Saffir-Simpson)",
    values = SSHS_COLOR_LABEL
  ) +
  labs(
    title="Number of Hurricanes per Season",
    x="Year",
    y="Number"
  ))
```

### Category 5 Hurricanes
```{sql connection=conn, output.var='cat5'}
SELECT
  season, name, wind_max_kt * 1.15078
FROM hurricane_summary
WHERE season > 1900 AND saffir_simpson(wind_max_kt) LIKE 'Category 5'
ORDER BY season DESC
```

```{r echo=FALSE}
names(cat5) = c("Season", "Name", "Peak Wind (MPH)")
pander(widen(cat5))
```

## Accumulated Cyclone Energy
Although simple, the Saffir Simpson Hurricane Wind Scale is not a complete description of a hurricane.
By referring using a storm's peak SSHWS rating to categorize its strength, we do not get an accurate
view of how much energy was produced during the entire *lifetime* of a storm.

```{sql connection=conn, output.var='ace'}
SELECT
  season,
  sum(ace(path_intensity)) AS ace
FROM hurricane_summary
WHERE season >= 1975
  AND st_intersects(path, st_makeenvelope(-70, -10, 0, 90, 4269))
GROUP BY season
ORDER BY season ASC
```

```{r}
ace_ts <- ts(data=ace, start=1975, end=2016, frequency=1)
ace_trend <- tslm(ace ~ season, data=ace_ts)

ggplot(ace, aes(x=season, y=ace)) + geom_bar(stat='identity') +
  geom_abline(intercept=coef(ace_trend)[1], slope=coef(ace_trend)[2]) +
  labs(title="Accumulated Cyclone Energy - North Atlantic",
       x="Season",
       y="Accumulated Cyclone Energy")
       
```

## References