---
title: "A Brief Introduction to Hurricanes"
output:
  html_notebook: default
  html_document:
    highlight: null
    theme: null
  pdf_document: default
---

```{r, message=FALSE, warning=FALSE, echo=FALSE}
source('db_conn.R')
source('hurricanes.R')
conn <- hurr_db_connect()

library(dplyr)
library(leaflet)
library(ggplot2)
library(pander)
# library(rjson) Trash
```

```{r ggplot Default Theme, echo=FALSE}
theme_set(theme_bw() + theme(
    legend.position="bottom",
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(colour = "black")))
```

For as long as humans have settled the coast, hurricanes have been a menace to civilization:

> Within the last two or three years, people have begun to think that the islands
  and peninsulas along the Texas and Louisiana Coast are unsafe for human abiding places. And
  Galveston Island is but a waif of the ocean, liable at any moment of
  being engulfed and submerged by the self-same power that gave it form.
    -- Braman's Information About Texas, 1858

## 1900 Galveston Hurricane
Surprisingly, with all the talk of the increasing risk caused by climate change, the deadliest storm in 
American history occured over a hundred years ago. 

### Hurricane Naming
Despite being the deadliest hurricane in American history, this storm is simply known by the location it struck. Meterologists did not decide to give official names were not given to hurricanes until 1953. Since then, names are recycled from six lists--each corresponding to different regions of the world. However, names associated with notoriously deadly or costly storms are removed.

### Path
While meteorologists had eyes on the 1900 hurricane as early as August 30, they were unable to get a good idea of its trajectory.
As shown on the map below, as it moved over the Gulf of Mexico over the Florida Keys, it rapidly strengthened and
then took a sudden westward turn. Although ships in the area were
able to get a fix on its trajectory, the limited communications abilities of the time prevented this information from reaching shore.

```{sql connection=conn, output.var='galv1900'}
WITH galv1900 AS
(SELECT path_intensity FROM hurricane_summary 
WHERE season::int = 1900
AND st_intersects(path, (SELECT geom FROM us_states WHERE name LIKE 'Texas')))
SELECT *
FROM json_to_recordset(array_to_json((SELECT * FROM galv1900)))
AS x(lon float, lat float, wind_max_kt float, time text, sshs text)
```

```{r, echo=FALSE}
# See: https://stackoverflow.com/questions/42989040/how-to-troubleshoot-missing-osm-tiles-in-leaflet-html-widget
m <- leaflet()
m <- addProviderTiles(m, providers$Esri.WorldImagery, group = "Esri")
for (i in 1:nrow(galv1900)) {
  m <- addCircleMarkers(m,
                        label=galv1900$time[i],
                        lng=galv1900$lon[i],
                        lat=galv1900$lat[i],
                        color=sshs_color(galv1900$sshs[i]))
}
m <- addLegend(m, position = c("bottomright"), title="Intensity",
               labels=SSHS_LABELS, colors=SSHS_COLOR_SCALE)
m
```

### The Aftermath
Before the storm, Galveston, Texas was a bustling coastal town. However, when it
struck on September 8, 1900 the 
storm with its 140 mph winds and massive storm surge killed between 6000-12000
people (out of a population of 37000). This effectively ended the "Golden Era of Galveston".

## Hurricane Tracking
At least in the United States, it is unlikely that another 1900 Galveston could occur.
Armed with a bevy of aircraft, radar, sallites, and computer models, today's metereologists 
are much more able to accurately predict the path of a hurricane and warn those in harms way.

### Effectiveness
The graph below speaks for itself in regards to the efficacy of modern hurricane tracking.

```{sql connection=conn, output.var='hurr_by_year'}
SELECT
  season, count(name), round(AVG(wind_max_kt)::numeric, 1),        
  saffir_simpson(AVG(wind_max_kt))
FROM hurricane_summary
WHERE season > 1900 AND wind_max_kt::int != -999
GROUP BY season
ORDER BY season
```

```{r "Hurricanes per Season", echo=FALSE}
add_notes(ggplot(hurr_by_year, aes(x=season, y=count, fill=saffir_simpson)) +
  geom_bar(stat='identity') +
  
  # Rescale to fit text annotations
  scale_y_continuous(limits=c(0, 160)) + 
  
  # Manually set legend for hurricane intensity
  scale_fill_manual(
    "Average Intensity",
    values = c("Tropical Depression" = "#5ebaff",
               "Tropical Storm" = "#00faf4",
               "Category 1" = "#ffffcc")
  ) +
  labs(
    title="Number of Hurricanes per Season",
    x="Year",
    y="Number"
  ))
```

### Motivation
The Weather Bureau (now the National Weather Service) didn't always have the blessing of adequate funding. However, 
Hurricanes Carol and Edna in 1954 reversed their fortunates, convincing Congress to
allocate more funds for hurricane research.

```{sql connection=conn, output.var='carol1954'}
SELECT *, saffir_simpson_color(wind_max_kt) as sshs_color
FROM hurricane_path('CAROL', 1954)
```

```{sql connection=conn, output.var='edna1954'}
SELECT *, saffir_simpson_color(wind_max_kt) as sshs_color
FROM hurricane_path('EDNA', 1954)
```

```{r, echo=FALSE}
m <- leaflet()
m <- addProviderTiles(m, providers$Esri.WorldImagery, group = "Esri")

# Hurricane Carol
m <- addPolylines(m,
                  lng=carol1954$lon,
                  lat=carol1954$lat,
                  color="white",
                  label="Hurricane Carol (1954)")
m <- addCircleMarkers(m,
                  label=carol1954$time,
                  lng=carol1954$lon,
                  lat=carol1954$lat,
                  color=carol1954$sshs_color)

# Hurricane Edna
m <- addPolylines(m, lng=edna1954$lon, lat=edna1954$lat,
                  label="Hurricane Edna (1954)", color="white")
m <- addCircleMarkers(m,
                  label=edna1954$time,
                  lng=edna1954$lon,
                  lat=edna1954$lat,
                  color=edna1954$sshs_color)

m <- addLegend(m, position = c("bottomright"), title="Intensity",
               labels=SSHS_LABELS, colors=SSHS_COLOR_SCALE)
m
```

### Aircraft
<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">Video of a flight through the eye of <a href="https://twitter.com/hashtag/Irma?src=hash&amp;ref_src=twsrc%5Etfw">#Irma</a> on <a href="https://twitter.com/hashtag/NOAA42?src=hash&amp;ref_src=twsrc%5Etfw">#NOAA42</a>. Flights on both the WP-3D Orion and G-IV <a href="https://twitter.com/hashtag/NOAA49?src=hash&amp;ref_src=twsrc%5Etfw">#NOAA49</a> continue. Credit Nick Underwood/NOAA <a href="https://t.co/9ini4bOnYF">pic.twitter.com/9ini4bOnYF</a></p>&mdash; NOAAHurricaneHunters (@NOAA_HurrHunter) <a href="https://twitter.com/NOAA_HurrHunter/status/905184657431506945?ref_src=twsrc%5Etfw">September 5, 2017</a></blockquote>
<script async src="//platform.twitter.com/widgets.js" charset="utf-8"></script>

### Radar

## References